---
title: "Project 1"
author: "Victor Alanis"
date: "6/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading the data
First, we are gonna read the data of the "activity.zip", and after that we are gonna make some small changes, i.e: clean the data and change the format.


```{r }
activity<- read.table(unz("activity.zip", "activity.csv"), header=T, sep=",")

activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")
clean <- activity[!is.na(activity$steps),]
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum )
colnames(sumTable)<- c("Date", "Steps")

```

## Histogram of the steps
```{r}
hist(sumTable$Steps,main= "histogram of steps per day",xlab="Steps")
```

##Mean and median
```{r}
as.integer(mean(sumTable$Steps))
as.integer(median(sumTable$Steps))

```

##Plot of the average daily activity pattern
```{r}
library(plyr)
library(ggplot2)
##pulling data without nas
clean <- activity[!is.na(activity$steps),]

##create average number of steps per interval
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))

##Create line plot of average number of steps per interval
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")

```

##The 5-minute interval that, on average, contains the maximum number of steps
```{r}
##Maximum steps by interval
maxSteps <- max(intervalTable$Avg)
##Which interval contains the maximum average number of steps
intervalTable[intervalTable$Avg==maxSteps,1]
```

##Code to describe and show a strategy for imputing missing data
```{r}
## Create the average number of steps per weekday and interval
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))

## Create dataset with all NAs for substitution
nadata<- activity[is.na(activity$steps),]
## Merge NA data with average weekday interval for substitution
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
## Reorder the new substituded data in the same format as clean data set
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")

##Merge the NA averages and non NA data together
mergeData <- rbind(clean, newdata2)
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")
```

##Histogram of the total number of steps taken each day after missing values are imputed
```{r}
## Creating the histogram of total steps per day, categorized by data set to show impact
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Black")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )
```

##Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
mergeData$DayCategory <- ifelse(mergeData$day %in% c("sabado", "domingo"), "Weekend", "Weekday")
library(lattice) 
## Summarize data by interval and type of day
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

##Plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```

